---
title: "R Markdown Final Individual Project"
author: "Ethan Sharp"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT



# BACKGROUND
 does higher education lead to a higher likelyhood of heart attack
 
 Higher education leads to a higher chance of heart attack because of the amount of stress


# METHODS
we will use a chi squared, binomial linear regression, and ANOVA


# DISCUSSION

## Interpretation - fill in analysis
# Example: Scatterplot with regression line

# (Inspired by R Graph Gallery scatter plot example)

# Load packages

library(ggplot2)

# Example dataset (replace this with your real CSV import)

# health_data <- read.csv("your_data.csv")

# Simulated example data for demonstration

set.seed(123)
health_data <- data.frame(
education_years = sample(8:20, 100, replace = TRUE),  # years of education
heart_attack = rbinom(100, 1, prob = 0.1 + 0.02 * (8:20)[sample(1:13, 100, replace = TRUE)] / 20)
)

# Plot: relationship between education level and heart attack rate

```{r}
# Example: Scatterplot with regression line

# (Inspired by R Graph Gallery scatter plot example)

# Load packages

library(ggplot2)

# Example dataset (replace this with your real CSV import)

# health_data <- read.csv("your_data.csv")

# Simulated example data for demonstration

set.seed(123)
health_data <- data.frame(
education_years = sample(8:20, 100, replace = TRUE),  # years of education
heart_attack = rbinom(100, 1, prob = 0.1 + 0.02 * (8:20)[sample(1:13, 100, replace = TRUE)] / 20)
)

# Plot: relationship between education level and heart attack rate

ggplot(health_data, aes(x = education_years, y = heart_attack)) +
geom_jitter(width = 0.2, height = 0.02, alpha = 0.6, color = "darkred") +
geom_smooth(method = "glm", method.args = list(family = "binomial"), se = TRUE, color = "black") +
labs(
title = "Relationship Between Education Level and Heart Attack Likelihood",
x = "Years of Education",
y = "Heart Attack Occurrence (0 = No, 1 = Yes)"
) +
theme_minimal(base_size = 13)

```

# REFERENCES     




2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.



