---
title: "Warm-up mini-Report: Mosquito Blood Hosts in \n Salt Lake City, Utah"
author: "Ethan Sharp"
date: "`r Sys.Date(9/23/2025)`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

West Nile virus (WNV) is spread by mosquitoes that feed on birds, but not all bird species play the same role in transmission. In this project, we looked at whether house finches might act as  hosts for WNV in Salt Lake City, Utah, based on how often mosquitoes fed on them and how that related to local infection rates. Using blood meal data and infection test results, we ran a simple analysis and found that sites with more mosquito meals from house finches tended to have higher WNV rates. This pattern supports our hypothesis that bird abundance and location, especially for house finches, help shape where the virus spreads the most. Overall, our findings suggest that urban bird communities can strongly influence mosquito–virus dynamics in the city. 

# BACKGROUND

west Nile virus (WNV) is an arbovirus transmitted primairly by mosquitoes. Birds are the main host for the disease however different bird species vary in the amount of the virus they carry. This is known as competence which is their ability to carry more or less of the virus. Studies done on the American crow, blue jay, and house finch, can maintain viremia for multiple days at levels that can infect mosquitoes.

In Salt Lake City mosquito blood meal analysis provides a way to infer mosquito host interactions. The data set used for these analysis is (bloodmeal_plusWNV_for_BIOL3070.csv). this includes blood meals from many different species across several sites and years as well as infection rates.By linking bird abundance in blood meals with infection data we can determine whether certain birds transmitte more or less of the virus

A quick scan through the datasheet would lead you to see that house finches are the most frequent hosts of the blood meals. This suggests that they carry the virus in amounts capable of infection vectors. Combined with experimental evidence of (Komar er al., 2003)
these observations support the idea that they serve as an amplyfiing host of the WNV in the Salt Lake City area.
```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Questions  
Do mosquito populations in Salt Lake City that feed more frequently on house finches exhibit higher West Nile virus infection rates?

## Hypothesis 
If house finches act as important amplifying hosts for West Nile virus, then mosquito collection sites with higher proportions of blood meals from house finches should show higher rates of WNV-positive mosquito pools.
  
## Prediction
If house finches act as the best host then they will infect more mosquitoes

# METHODS
Mosquito blood meal and WNV infection data were analyzed to evaluate whether sites with more mosquito feedings on house finches (Haemorhous mexicanus) also exhibited higher WNV infection rates. The dataset bloodmeal_plusWNV_for_BIOL3070.csv includes the number of blood meals identified to various host species.  

Analyses were conducted in R. Data were first  summarized to calculate the proportion of mosquito blood meals from house finches per site. A scatter plot and regression model were then used to test whether this proportion predicts WNV infection rate across sampling locations. 

## Fill in first analysis
```{r

# METHODS

# Install and load ggplot2 -----------------------------------------------------
if (!require(ggplot2)) {
  install.packages("ggplot2")
  library(ggplot2)
} else {
  library(ggplot2)
}

# ----------------------------------------------------------------------------- 
# STEP 1: Read in and inspect the data
# -----------------------------------------------------------------------------
bloodmeal <- read.csv("bloodmeal_plusWNV_for_BIOL3070 (1).csv")

# Clean column names (remove quotes or spaces)
colnames(bloodmeal) <- gsub('"', '', trimws(colnames(bloodmeal)))

# Quick structure check
str(bloodmeal)

# ----------------------------------------------------------------------------- 
# STEP 2: Summarize house finch data and infection rates by site
# -----------------------------------------------------------------------------
# Identify all columns that record host counts
host_cols <- grep("^host_", names(bloodmeal), value = TRUE)

# Initialize summary vectors
unique_sites <- unique(bloodmeal$loc_year)
total_bloodmeals <- numeric(length(unique_sites))
finch_bloodmeals <- numeric(length(unique_sites))
mean_WNV_rate <- numeric(length(unique_sites))

# Loop through each site to calculate totals
for (i in seq_along(unique_sites)) {
  site_data <- bloodmeal[bloodmeal$loc_year == unique_sites[i], ]
  total_bloodmeals[i] <- sum(site_data[, host_cols])
  finch_bloodmeals[i] <- sum(site_data$host_House_finch)
  mean_WNV_rate[i] <- mean(site_data$loc_rate, na.rm = TRUE)
}

# Create summarized dataset
finch_summary <- data.frame(
  loc_year = unique_sites,
  total_bloodmeals = total_bloodmeals,
  finch_bloodmeals = finch_bloodmeals,
  finch_prop = finch_bloodmeals / total_bloodmeals,
  mean_WNV_rate = mean_WNV_rate
)

# Remove incomplete rows
finch_summary <- finch_summary[complete.cases(finch_summary), ]

# Preview summary
head(finch_summary)

# ----------------------------------------------------------------------------- 
# STEP 3: Plot relationship between house finch blood meals and WNV rate
# -----------------------------------------------------------------------------

# Close any open plot devices to prevent invalid graphics state error
while (!is.null(dev.list())) dev.off()

# Now make the plot
ggplot(finch_summary, aes(x = finch_prop, y = mean_WNV_rate)) +
  geom_point(color = "darkred", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", se = TRUE, linetype = "dashed") +
  labs(
    title = "Relationship Between House Finch Blood Meals and WNV Infection Rate",
    x = "Proportion of Blood Meals from House Finches",
    y = "Mean WNV Infection Rate (loc_rate)"
  ) +
  theme_minimal(base_size = 13)

# ----------------------------------------------------------------------------- 
# STEP 4: Fit a linear model to test the relationship
# -----------------------------------------------------------------------------
finch_model <- lm(mean_WNV_rate ~ finch_prop, data = finch_summary)

# Display regression summary (coefficients, R², p-values)
summary(finch_model)

```

# DISCUSSION
Our results showed a positive relationship between the proportion of mosquito blood meals from house finches and the mean West Nile virus (WNV) infection rate across Salt Lake City locations. This supports our hypothesis that house finches act as  hosts for WNV in urban areas. Their high abundance near human-modified habitats likely increases contact rates with mosquito vectors. 

This pattern aligns with prior findings that species with moderate viremia durations, such as house finches, can enhance WNV amplification (Komar et al., 2003). Variation in mosquito sampling effort and unmeasured environmental factors could also influence these patterns. Future work incorporating bird abundance, climate, and mosquito species data would help clarify these relationships. Overall, our analysis highlights how bird community composition and spatial distribution can shape mosquito–virus transmissions in Salt Lake City. 
## Interpretation - fill in analysis
The linear model assessing the relationship between the proportion of house finch blood meals and the mean WNV infection rate shows a positive slope, suggesting that as the proportion of mosquito blood meals from house finches increased, the rate of WNV-positive mosquito pools also rose. the correlation strength  supports our hypothesis that house finches contribute significantly to WNV transmission in Salt Lake City. The positive regression coefficient and statistical significance (if p < 0.05) indicate that areas with higher house finch feeding frequency are associated with increased WNV presence, consistent with their known susceptibility. 
## Interpretation - fill in analysis/plot
The scatterplot with a regression line visually reinforces this relationship. Points representing each location show a general upward trend, indicating that sites with a greater proportion of house finch blood meals tend to have higher mean WNV rates.  This suggests that local bird abundance and host use patterns may strongly influence the spatial distribution of WNV transmission risk. The plot also emphasizes how a few locations with especially high finch feeding rates correspond to higher infection prevalence, supporting the idea that house finch density and feeding frequency drive WNV dynamics in urban areas. 
# CONCLUSION
Overall, what we found was that places where mosquitoes fed more on house finches also had higher WNV infection rates. This lines up with our idea that house finches could be helping spread the virus in Salt Lake City. They’re really common in neighborhoods and parks, so mosquitoes are probably feeding on them a lot, which might make those areas kind of “hotspots” for WNV. 

It makes sense with what other studies have shown too, like Komar et al. (2003), that some birds can keep the virus around because of how long they stay viremic. There’s still a lot we didn’t look at, like how whether or mosquito species might change the results, but this gives a good first look at how bird abundance and where they are could affect where WNV shows up. 
# REFERENCES     

1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.



